### 2016-07-25 finalText/abstract.tex
Final Version of the Text and Short Paper

Esta dissertação tem como objetivo a implementação em hardware, numa plataforma
FPGA, de uma Rede Neuronal \textit{Long Short-Term Memory} (LSTM).

As redes neuronais são uma das técnicas mais usadas na área da Aprendizagem Computacional Profunda para tarefas em
que a programação explícita de \textit{software}, para além de ser impraticavelmente complexa, é necessário que haja uma qualidade
de tomada de decisão semelhante ou superior à inteligência Humana. Esta rede em particular, a LSTM, é uma rede recursiva, na medida em que a saída
da cada neurónio, num instante de tempo, \emph{também} serve de entrada no instante seguinte, e assim
os elementos de memória nela presentes conseguem encontrar padrões também em sequências de dados,
apresentando vantagens claras, neste campo, face a redes neuronais convencionais.

As aplicações desta rede são inúmeras, como se poderá atestar no capítulo
que faz o levantamento do estado-da-arte. Embora a implementação em \textit{software}
seja a solução comum para todas estas arquitecturas, as implementações em \textit{hardware}
são ainda poucas e os benefícios do paralelismo inerente a uma plataforma de \textit{hardware} dedicado (no caso deste trabalho, uma FPGA) não
são aproveitados. É nesse enquadramento que este trabalho se posiciona, apresentando uma implementação inédita de LSTM em FPGA, fazendo apenas uso de 
recursos internos à mesma, que tem uma melhoria de rapidez de processamento de cerca de 195 vezes face 
a uma implementação \textit{software}, assim como a proposta de um circuito que permita fazer o 
treino \textit{on-chip} da rede, usando um método de perturbações estocásticas simultâneas (SPSA).

@@

Neural Networks are one of the most commonly used techniques in Deep Learning for tasks
where explicit programming of software, besides impractically complex, it is necessary that
the quality of the decision making is similar, or even better, than that of Human intelligence.
This particular type of network, LSTM, is a recursive network, given that the neuron outputs in a 
given time is \emph{also} fed as the input in the next time instant, and that way their memory elements can
make sense of patterns within data sequences, unlike conventional neural networks.

This type of network, as well as many other kinds of networks, have been profusely implemented
in software, and their practical applications are plentiful, as one can attest
in the State of the Art chapter. However, the benefits of the inherent parallelism offered by a dedicated hardware platform
(in this work, an FPGA) are not availed, and there are relatively few the implementation of Machine Learning algorithms in
these kind of platforms. It is within this grounding that this work positions itself, implementing an LSTM Network in FPGA, using only its internal resources, and achieving an $\times195$ increase in processing time when compared to a software implementation, and also proposes a circuit that will allow on-chip training for this network, using simultaneous stochastic perturbations (SPSA).

>>>>

Esta dissertação tem como objetivo a implementação em FPGA de uma Rede Neuronal \textit{Long Short-Term Memory} (LSTM).

As redes neuronais são uma das técnicas mais usadas na área da Aprendizagem Computacional Profunda.
Esta rede em particular, a LSTM, é uma rede recursiva, na medida em que a saída
da cada neurónio, num instante de tempo, \emph{também} serve de entrada no instante de tempo seguinte, e
os seus elementos de memória conseguem recordar padrões em sequências de dados,
apresentando vantagens claras, neste campo, face a redes neuronais convencionais.

As aplicações das Redes Neuronais são inúmeras, como se poderá atestar no capítulo
que faz o levantamento do estado-da-arte. Embora a implementação em \textit{software}
seja a solução comum para todas estas arquitecturas, as implementações em \textit{hardware}
são ainda poucas e os benefícios do paralelismo inerente a uma plataforma de \textit{hardware} dedicado não
são aproveitados. É nesse enquadramento que este trabalho se posiciona, apresentando uma implementação inédita de LSTM em FPGA,
fazendo apenas uso de recursos internos à mesma, apresentando um \textit{speed-up} de 251 vezes face
a uma implementação \textit{software} --- executada num Computador de alta \textit{performance} --- assim como a
proposta de um circuito auxiliar que permita fazer o treino \textit{on-chip} da rede, usando um método de perturbações
estocásticas simultâneas (SPSA).

@@

Neural Networks are one of the most commonly used techniques in Deep Learning.
This particular type of network, as LSTM Network, is a recursive network, given that the neuron outputs in a
given time are \emph{also} fed as the input in the next time step and, that way, their memory elements can
make sense of patterns within data sequences, unlike conventional neural networks.

This type of network, as well as many other kinds of networks, have been profusely implemented
in software, and their practical applications are plentiful, as one can attest
in the State of the Art chapter. However, there are only few FPGA implementation of these algorithms, and the benefits of the inherent
parallelism offered by a dedicated hardware platform are not availed. This work, then, implements an LSTM Network in FPGA, using only
its internal resources, and achieving a $251 \times$ \textit{speed-up} in processing time when compared to a software implementation,
running on a high-end Desktop. A circuit that will allow on-chip training for this network, using simultaneous stochastic perturbations (SPSA),
is also proposed.

### 2016-07-25 finalText/acknows.tex
Final Version of the Text and Short Paper

Agradeço a todos os professores que se cruzaram comigo na minha vida e que, através da sua ação
me transmitiram os conhecimentos necessários para me formar como Engenheiro e produzir este trabalho,
que espero que seja de uso para a comunidade científica e para a Humanidade. Uma palavra especial
ao meu orientador deste trabalho, o Prof. João Canas Ferreira, que desde o primeiro momento acolheu
esta proposta de trabalho e soube ver nela o seu potencial devido, e ao co-orientador, o Ivo Timóteo
que mais do que minha referência para a parte teórica deste trabalho, é um bom amigo.

@@

Uma última palavra de apreço para as pessoas especiais da minha vida, os meus pais, que me formaram como Pessoa e me
incutiram força em todos os momentos da vida, do qual esta tese não é excepção, e à minha namorada, a Mafalda, que sempre 
se entusiasmou e interessou por todos os pequenos progressos que fui atingindo ao longo destes meses.

>>>>

Agradeço a todos os professores que se cruzaram comigo na minha vida e que, através da sua ação,
me transmitiram os conhecimentos necessários para me formar como Engenheiro e produzir este trabalho,
que espero que seja de uso para a comunidade científica e para a Humanidade. Uma palavra especial
ao meu orientador deste trabalho, o Prof. João Canas Ferreira, que desde o primeiro momento acolheu
esta proposta de trabalho e soube ver nela o seu potencial devido, e ao co-orientador, o Ivo Timóteo,
que mais do que minha referência para a parte teórica deste trabalho, é um bom amigo.

@@

Uma última palavra de apreço para as pessoas especiais da minha vida. Os meus pais, que me formaram como Pessoa e me
incutiram força em todos os momentos da vida, do qual esta tese não é excepção, e à minha namorada, a Mafalda, que apesar
de não ser da minha área, sempre se entusiasmou e interessou por todos os progressos que fui atingindo ao longo destes meses.

### 2016-07-25 finalText/intro.tex
Final Version of the Text and Short Paper

In this work, a hardware implementation of an LSTM Neural Network is presented, along with a proposal for a digital circuit that performs on-chip training. In the reference example, the proposed hardware network was 195 times
faster than the reference software implementation. The training convergence, although simulated, is not conclusive.

\section{Background}\label{sec:intro_back}
Artificial Neural Networks are one of the most popular models in the field of Machine Learning. As the name suggests, their operation is inspired by the operation of the building blocks of our brain, the \textbf{neurons}. In spite of its high performance, one of their shortcomings is the fact that they cannot retain temporal dependences among incoming data samples, thus not being suitable to process time-series data, such as audio, video or other kinds of time-varying data streams, where current inputs have a high temporal dependence with previous and future inputs.

To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail to recall temporal dependences that extend over an large period of time, for the reasons that we will understand in Sections~\ref{sec:theorBack_rnn} and~\ref{sec:theorBack_lstm}.Thus, in 1997 Hochreiter et al. proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and the input/output/read/write access is controlled by individual gates that are activated both by the incoming data samples, but also by the outputs from the previous time-step (it is an RNN after all). They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.

\section{Motivation}\label{sec:intro_mot}
Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be sped up by a dedicated hardware structure.

@@

\section{Objectives}\label{sec:intro_obj}
Taking into account the considerations done in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}. Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate some of the applications portrayed in~\ref{sec:sa_apps}. Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.

\section{People Involved}\label{sec:intro_people}
Besides me, the candidate to the Master's Degree, there are two more people involved, namely

@@

\section{Overview of the Document}\label{sec:intro_overview}
Chapter~\ref{chap:theorBack}, the theoretical foundations will be layed out. Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs. Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final solution, which will be outlined in Section~\ref{chap:propSol}.

In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.

@@

\textbf{Disclaimer}: since this is a very innovative work, I will not present here any Verilog source code to protect my Intellectual Property. However, you can ask permission to have it by \href{mailto:ee11126@fe.up.pt}{emailing me}.

>>>>

In this work, a hardware implementation of an LSTM Neural Network is presented, along with a proposal for a digital circuit that performs on-chip training.
In the reference example, the proposed hardware network was 251 times faster than the reference software implementation running in a high-end Desktop.
The training convergence, although simulated, was not conclusive.

\section{Background}\label{sec:intro_back}
Artificial Neural Networks are one of the most popular algorithms in the field of Machine Learning. As the name suggests, their operation is inspired by
the operation of the building blocks of our brain, the \textbf{neurons}. In spite of its high performance, one of their shortcomings is the fact that
they cannot retain temporal dependences in a stream of data samples, thus not being suitable to process time-series data, such as audio, video or other
kinds of time-varying data streams, where inputs at a given time step have a high temporal dependence with previous and future inputs.

To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail
to recall temporal dependences that extend over a large period of time, for the reasons that we will understand in Sections~\ref{sec:theorBack_rnn} and~\ref{sec:theorBack_lstm}.
In 1997 Hochreiter et al. proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and
the input/output/read/write access is controlled by individual gates that are activated both by the incoming data samples, but also by the outputs
from the previous time-step (it is an RNN after all). They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.

\section{Motivation}\label{sec:intro_mot}
Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based. But the parallel nature of the
structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that has only recently been done
once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve
software solution, it makes use of external memory to store the network parameters, which undermines the network performance. Furthermore, the
design is not flexible to accommodate networks of different sizes, and lacks the ability to perform on-chip learning.

@@

\section{Objectives}\label{sec:intro_obj}
Taking into account the considerations done in~\ref{sec:intro_mot}, a hardware implementation of an
LSTM Network is proposed, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
Moreover, the HDL description of the network is parameterizable, meaning that to synthesize a network for a given size, we only need to
change the appropriate parameter at synthesis time, therefore requiring no redesign effort whatsoever, unlike~\cite{Chang15}.

\section{People Involved}\label{sec:intro_people}
Besides myself, the candidate to the Master's Degree, there are two more people involved, namely

@@

\section{Overview of the Document}\label{sec:intro_overview}
In Chapter~\ref{chap:theorBack}, the theoretical foundations will be laid out. Section~\ref{sec:tb_basicCon} presents the
basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final
solution, which will be outlined in Section~\ref{sec:train-circ}.

In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications (Section~\ref{sec:sa_apps}), their
hardware implementations (Section~\ref{sec:sa_hardware}) and the current work regarding the chosen training algorithm (Section~\ref{sec:sa_training})
are presented.

@@
